---
title: "Download and prepare taxonomy files (trees and GTDB taxonomies)"
author: "Giacomo Antonello"
date: "2025-12-29"
output: 
  html_document:
    toc: true
---

```{r setup, message=FALSE, warning=FALSE}
library(zen4R) # Zenodo API for R
library(knitr) 
library(tools) # md5sum
library(ape)   # read.tree, write.tree, Ntip

opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)

mpa_doi <- "https://doi.org/10.5281/zenodo.3965462"

outdir <- "../tmp/metaphlan_downloads"
dir.create(outdir, recursive = TRUE)
```

# Download latest standard taxonomy from `cmprod1`

The latest MetaPhlAn database name is retrieved from the `mpa_latest` pointer file on
the CIBIO server. The corresponding species taxonomy file maps each SGB identifier to its
full MetaPhlAn taxonomy string and is stored compressed in `inst/extdata`.

```{r}
cmprod_url <- "http://cmprod1.cibio.unitn.it/biobakery4/metaphlan_databases"
mpa_latest <- readLines(sprintf("%s/mpa_latest", cmprod_url))

download.file(
  url      = sprintf("%s/%s_species.txt.bz2", cmprod_url, mpa_latest),
  destfile = file.path("..", "inst", "extdata", paste0(mpa_latest, "_speciesTaxonomy.tsv.bz2"))
)

latest_mpa_taxonomy <- read.delim(
  file.path("..", "inst", "extdata", paste0(mpa_latest, "_speciesTaxonomy.tsv.bz2")),
  header    = FALSE,
  col.names = c("SGB", "Taxonomy")
)
```

# List all Zenodo entries of MetaPhlAn

All public versions of the MetaPhlAn database are archived on Zenodo. We retrieve
the full version history, drop development releases (beta/test), deduplicate by DOI
(keeping only the latest edit of each release), and sort newest-first.

```{r}
zenodo_versions_with_doi <- get_versions(mpa_doi)

zenodo_versions_with_doi <- zenodo_versions_with_doi[, c("date", "version", "doi")]
zenodo_versions_with_doi <- zenodo_versions_with_doi[!grepl("beta|test", zenodo_versions_with_doi$version), ]
zenodo_versions_with_doi <- zenodo_versions_with_doi[!duplicated(zenodo_versions_with_doi$doi), ]
zenodo_versions_with_doi <- zenodo_versions_with_doi[rev(order(zenodo_versions_with_doi$date)), ]
rownames(zenodo_versions_with_doi) <- seq_len(nrow(zenodo_versions_with_doi))

kable(zenodo_versions_with_doi)
```

# Download and unzip all versions from 3.0 to the latest (~3 min)

Downloads are throttled (30 s pause every 8 requests) to stay within Zenodo's rate
limits. Each zip is extracted into a same-named subdirectory of `outdir`.

```{r, cache=TRUE, message=FALSE, eval=!interactive()}
for (i in seq_len(nrow(zenodo_versions_with_doi))) {
  if (i %% 8 == 0) Sys.sleep(30)
  download_zenodo(doi = zenodo_versions_with_doi$doi[i], path = outdir)
}

invisible(sapply(
  list.files(outdir, pattern = "\\.zip$", full.names = TRUE),
  function(x) unzip(zipfile = x, overwrite = TRUE, exdir = gsub("\\.zip$", "", x))
))
```

# Download tree exceptions from `cmprod`

Two trees are not available via Zenodo and must be fetched directly from the CIBIO
server: the corrected v3.1 tree (the Zenodo archive ships the v3.0 tree for that
release) and the latest v4.2+ tree.

```{r, warning=FALSE}
download.file(
  "http://cmprod1.cibio.unitn.it/biobakery3/mpa_v31_CHOCOPhlAn_201901_species_tree.nwk",
  destfile = file.path(outdir, "MetaPhlAn-3.1.0", "mpa_v31_CHOCOPhlAn_201901_species_tree.nwk")
)

download.file(
  "http://cmprod1.cibio.unitn.it/biobakery4/metaphlan_databases/mpa_vJan25_CHOCOPhlAnSGB_202503.nwk",
  destfile = file.path(outdir, "MetaPhlAn-4.2.0", "mpa_vJan25_CHOCOPhlAnSGB_202503.nwk")
)
```

# Find all `.nwk` tree files and build a directory table

```{r}
trees.vec        <- list.files(outdir, pattern = "\\.nwk$", recursive = TRUE, full.names = TRUE)
trees.md5sum     <- md5sum(trees.vec)
trees.split.list <- strsplit(trees.vec, "/", fixed = TRUE)
trees.list       <- lapply(trees.vec, read.tree)
names(trees.list) <- basename(trees.vec)

trees.df <- data.frame(
  filePath          = trees.vec,
  tree_filename     = basename(trees.vec),
  MetaPhlAn_version = sapply(trees.split.list, function(x) gsub("MetaPhlAn-", "", x[4], fixed = TRUE)),
  tree_Ntips        = sapply(trees.list, Ntip),
  md5sum            = trees.md5sum,
  row.names         = NULL
)
```

Remove the incorrect v3.1.0 tree (Zenodo ships the v3.0 tree for that release; the
correct one was downloaded above from `cmprod`).

```{r}
trees.df <- trees.df[
  !(trees.df$MetaPhlAn_version == "3.1.0" &
    trees.df$tree_filename == "mpa_v30_CHOCOPhlAn_201901_species_tree.nwk"), ]
```

# Extend tip labels with full taxonomy and write trees as `.nwk.bz2`

Older trees store only bare SGB numbers as tip labels (no `k__`/`p__` prefixes).
For those trees we look up the full MetaPhlAn taxonomy string from `latest_mpa_taxonomy`
and prepend it, giving every tip label the form:
`k__Bacteria|...|s__Bacteroides fragilis|t__SGB1234`

All trees are then written as bzip2-compressed Newick files into `inst/extdata/trees/`.

```{r}
destdir <- "../inst/extdata/trees"
dir.create(destdir, recursive = TRUE)

# Write each tree as a bzip2-compressed Newick file
invisible(lapply(names(trees.list), function(filename) {
  out_path <- sprintf("%s/%s.bz2", destdir, filename)
  write.tree(phy = trees.list[[filename]], file = bzfile(out_path))
}))

# Write the directory/manifest table (plain TSV, no row names)
write.table(
  trees.df[!duplicated(trees.df$md5sum),][, -1],   # drop filePath — not meaningful after install
  file      = file.path(destdir, "trees_files_directory.tsv"),
  sep       = "\t",
  row.names = FALSE,
  col.names = TRUE,
  quote     = FALSE
)
```

# Format and write GTDB taxonomy tables as `.tsv.bz2`

GTDB files are identified by the `*GTDB*` glob across all downloaded version folders.
Duplicates (same md5) are dropped so each unique table is written only once.

The raw GTDB format is a two-column file: SGB | semicolon-delimited taxonomy string.
We split the taxonomy string into named columns (Domain … Species) and append the SGB
column, then write the result as a bzip2-compressed TSV.

```{r}
GTDBs.vec        <- list.files(outdir, pattern = ".*GTDB.*", recursive = TRUE, full.names = TRUE)
GTDBs.md5sum     <- md5sum(GTDBs.vec)
GTDBs.split.list <- strsplit(GTDBs.vec, "/")

GTDBs.df <- data.frame(
  filePath          = GTDBs.vec,
  GTDB_filename     = basename(GTDBs.vec),
  MetaPhlAn_version = sapply(GTDBs.split.list, function(x) gsub("MetaPhlAn-", "", x[4], fixed = TRUE)),
  md5sum            = GTDBs.md5sum
)

# Keep only one copy of each unique file (by md5)
GTDBs.df_unique <- GTDBs.df[!duplicated(GTDBs.df$md5sum), ]
```

```{r}
# Read each unique GTDB file using its full path (filePath column)
files_raw <- lapply(GTDBs.df_unique$filePath, read.delim,
                    col.names = c("SGB", "taxonomy_long"))
names(files_raw) <- GTDBs.df_unique$GTDB_filename

# Split the semicolon-delimited taxonomy string into one column per rank
files_clean <- lapply(files_raw, function(x) {
  ranks <- do.call(rbind, strsplit(x$taxonomy_long, ";"))
  df    <- cbind.data.frame(ranks, SGB = x$SGB, stringsAsFactors = FALSE)
  colnames(df) <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species", "SGB")
  df
})
```

```{r}
destdir <- "../inst/extdata/GTDB"
dir.create(destdir, recursive = TRUE)

# Write each table as a bzip2-compressed TSV
invisible(lapply(names(files_clean), function(filename) {
  out_path <- file.path(destdir, sub("\\.tsv$", ".tsv.bz2", filename))
  write.table(
    files_clean[[filename]],
    file      = bzfile(out_path),
    sep       = "\t",
    row.names = FALSE,
    quote     = FALSE
  )
}))

# Write the manifest (plain TSV — small file, no need to compress)
write.table(
  GTDBs.df_unique[, c("GTDB_filename", "md5sum")],
  file      = file.path(destdir, "GTDB_files_directory.tsv"),
  sep       = "\t",
  row.names = FALSE,
  quote     = FALSE
)
```

# sessionInfo

```{r}
sessionInfo()
```
