---
title: "Download and prepare taxonomy files (trees and standard_taxonomy taxonomies)"
author: "Giacomo Antonello"
date: "2025-12-29"
output: 
  html_document:
    toc: true
---

```{r setup, message=FALSE, warning=FALSE}
library(zen4R) # Zenodo API for R
library(knitr) 
library(tools) # to unzip downloaded files
library(ape) # to manipulated phylogenetic trees

opts_chunk$set(
  message = FALSE, 
  warning = FALSE
)

mpa_doi <- "https://doi.org/10.5281/zenodo.3965462"

outdir <- "../tmp/metaphlan_downloads"
dir.create(outdir, recursive = TRUE)
```

# Download latest standard taxonomy from `cmprod1`

```{r}
cmprod_url <- "http://cmprod1.cibio.unitn.it/biobakery4/metaphlan_databases"
mpa_latest <- readLines(sprintf("%s/mpa_latest", cmprod_url))

download.file(url = sprintf("%s/%s_species.txt.bz2", cmprod_url, mpa_latest), destfile = file.path("..", "inst", "extdata", paste0(mpa_latest, "_speciesTaxonomy.tsv.bz2")))

latest_mpa_taxonomy <- read.delim(file.path("..", "inst", "extdata", paste0(mpa_latest, "_speciesTaxonomy.tsv.bz2")), header = FALSE, col.names = c("SGB", "Taxonomy"))
```

# List all Zenodo entries of MetaPhlAn

```{r}
# download all version of the main metaphlan zenodo entry
zenodo_versions_with_doi <- get_versions(mpa_doi)

# subset only relevant dates, date entered and last modified are less relevant
zenodo_versions_with_doi <- zenodo_versions_with_doi[, c("date", "version", "doi")]

# remove beta and test versions
zenodo_versions_with_doi <- zenodo_versions_with_doi[!grepl("beta|test", zenodo_versions_with_doi$version),]

# remove duplicated rows, in case edits were made, but we only care 
# about the latest
zenodo_versions_with_doi <- zenodo_versions_with_doi[!duplicated(zenodo_versions_with_doi$doi),] 

# reorder from latest to oldest
zenodo_versions_with_doi <-  zenodo_versions_with_doi[rev(order(zenodo_versions_with_doi$date)), ]

rownames(zenodo_versions_with_doi) <- seq_len(nrow(zenodo_versions_with_doi))
kable(zenodo_versions_with_doi)
```

# Download and unzip all versions from 3.0 to the latest (\~ 3 min)

```{r, cache=TRUE, message=FALSE, eval=!interactive()}
# download sequentially
for(i in seq_len(nrow(zenodo_versions_with_doi))){
  if(i %% 8 == 0){
    Sys.sleep(30)
  }
  download_zenodo(doi = zenodo_versions_with_doi$doi[i], path = file.path(outdir))
}

# unzip in block
invisible(sapply(list.files(outdir, pattern = "*.zip", full.names = TRUE), function(x)
  unzip(
    zipfile = x,
    overwrite = TRUE,
    exdir = gsub(".zip", "", x)
  )))
```

# Download tree exception from cmprod into the 3.1 version

```{r, warning=FALSE}
# newick tree for metaphlan v 3.1.0:
invisible(download.file("http://cmprod1.cibio.unitn.it/biobakery3/mpa_v31_CHOCOPhlAn_201901_species_tree.nwk", destfile = file.path(outdir, "MetaPhlAn-3.1.0", "mpa_v31_CHOCOPhlAn_201901_species_tree.nwk")))

# newick tree for metaphlan v 4.2.0 and above
invisible(download.file("http://cmprod1.cibio.unitn.it/biobakery4/metaphlan_databases/mpa_vJan25_CHOCOPhlAnSGB_202503.nwk", destfile = file.path(outdir, "MetaPhlAn-4.2.0", "mpa_vJan25_CHOCOPhlAnSGB_202503.nwk")))
```

# Find all files with `.nwk` file extension

```{r}
trees.vec <- list.files(outdir, pattern = "*.nwk", recursive = TRUE, full.names = TRUE)
trees.md5sum <- md5sum(trees.vec)
trees.split.list <- strsplit(trees.vec, "/", fixed = TRUE)

# for each tree, find number of leaves/tips

trees.list <- lapply(trees.vec, read.tree)
names(trees.list) <- basename(trees.vec)

trees.df <- data.frame(
  filePath = trees.vec,
  tree_filename = basename(trees.vec), 
  MetaPhlAn_version = sapply(trees.split.list, function(x) gsub("MetaPhlAn-", "", x[4], fixed = TRUE)),
  tree_Ntips = sapply(trees.list,  Ntip),
  md5sum = trees.md5sum
  )

rownames(trees.df) <- NULL
```

Manually prune incorrect 3.1.0 tree, the correct one is downloaded from `cmprod`

```{r}
trees.df <- trees.df[!(trees.df$MetaPhlAn_version == "3.1.0" & trees.df$tree_filename == "mpa_v30_CHOCOPhlAn_201901_species_tree.nwk"),]
```

# Extend taxonomy and write trees in `.bz2` format into `inst/extdata`

```{r}
destdir <- "../inst/extdata/trees"
dir.create(destdir, recursive = TRUE)

# taxonomy as vector
latest_mpa_taxonomy.chr <- paste0(latest_mpa_taxonomy$Taxonomy, "|", "t__", latest_mpa_taxonomy$SGB)
names(latest_mpa_taxonomy.chr) <- paste0("t__", latest_mpa_taxonomy$SGB)

trees_improved_taxonomy.list <- lapply(trees.list, function(x) {
  if(!any(grepl("__", x$tip.label))){
    SGBLabels <- paste0("t__SGB", x$tip.label)
    x$tip.label <- latest_mpa_taxonomy.chr[SGBLabels]
    return(x)
  } else{
    return(x)
  }
}
)

for(pt in names(trees_improved_taxonomy.list)){
  write.tree(phy = trees_improved_taxonomy.list[[pt]], file = bzfile(sprintf("%s/%s.bz2", destdir, pt)))
}

# write all trees found in a map/directory file
write.table(trees.df[,-1], file = file.path(destdir, "trees_files_directory.tsv"), sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
```

# Format and write GTDB tables

Find all GTDB files

```{r}
GTDBs.vec <- list.files(outdir, pattern = "*GTDB.*", recursive = TRUE, full.names = TRUE)
GTDBs.md5sum <- md5sum(GTDBs.vec)

GTDBs.split.list <- strsplit(GTDBs.vec, "/")

GTDBs.df <- data.frame(
  filePath = GTDBs.vec,
  GTDB_filename = basename(GTDBs.vec), 
  MetaPhlAn_version = sapply(GTDBs.split.list, function(x) gsub("MetaPhlAn-", "", x[4], fixed = TRUE)),
  md5sum = GTDBs.md5sum
)

GTDBs.df_unique <- GTDBs.df[,c("GTDB_filename", "md5sum")]
GTDBs.df_unique <- GTDBs.df_unique[!duplicated(GTDBs.df_unique),]
```

Load and wrangle GTDB files

```{r}
files_raw <- lapply(rownames(GTDBs.df_unique), read.delim, col.names = c("SGB", "taxonomy_long"))
names(files_raw) <- GTDBs.df_unique$GTDB_filename

files_clean <- lapply(files_raw, function(x){
  x2 <- do.call(rbind, strsplit(x[[2]], ";"))
  x3 <- cbind.data.frame(x2, x[[1]])
  colnames(x3) <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species", "SGB")
  return(as.data.frame(x3))
})
```

Write GTDB files

```{r}
destdir <- "../inst/extdata/GTDB"

dir.create(destdir, recursive = TRUE)
invisible(lapply(names(files_clean), function(filename)
  write.table(
    files_clean[[filename]],
    file = file.path(
      destdir,
      filename),
      sep = "\t",
      row.names = FALSE
    )
  ))

write.table(x = GTDBs.df_unique, file = file.path(destdir, "GTDB_files_directory.tsv"), sep = "\t", row.names = FALSE)
```

# Save latest mpa version as variabl

```{r}
save(mpa_latest, file = file.path("..", "inst", "extdata", "mpa_latest.rda"))
```


# sessionInfo

```{r}
sessionInfo()
```
