---
title: "Basic package usage"
author: 
  - name: Giacomo Antonello
    affiliation: CUNY
package: ChocoPhlAnTaxonomies
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Basic package usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 4,
    warning = FALSE,
    message = FALSE
)
library(TreeSummarizedExperiment)
library(ChocoPhlAnTaxonomies)
```


# Validate a ChocoPhlAn version

In order to be able to map the wanted SGB to an available taxonomy file.
Validation is not tested, it was mostly to allow convergence of the 202403 tree 
into the 202307 one.

```{r}
validate_chocophlan_version("202403")
```

# Find SGB full taxonomy name by taxonomy type and version

```{r}
get_SGB_full_taxonomy(SGB = "SGB20358", taxonomy = "mpa", chocophlan_datestamp = 202403)
get_SGB_full_taxonomy(SGB = "SGB20358", taxonomy = "GTDB", chocophlan_datestamp = 202403)

get_SGB_full_taxonomy(SGB = "SGB20358", taxonomy = "mpa", chocophlan_datestamp = "latest")
get_SGB_full_taxonomy(SGB = "SGB20358", taxonomy = "GTDB", chocophlan_datestamp = "latest")

```

`t__` in front is also supported:

```{r}
get_SGB_full_taxonomy(SGB = "t__SGB20358", taxonomy = "mpa", chocophlan_datestamp = 202403)
get_SGB_full_taxonomy(SGB = "t__SGB20358", taxonomy = "GTDB", chocophlan_datestamp = 202403)

get_SGB_full_taxonomy(SGB = "t__SGB20358", taxonomy = "mpa", chocophlan_datestamp = "latest")
get_SGB_full_taxonomy(SGB = "t__SGB20358", taxonomy = "GTDB", chocophlan_datestamp = "latest")
```

# Load taxonomy tree into `TreeSummarizedExperiment` object

## Pre-2021 MetaPhlAn data

This is an edge case to show how the 

```{r}
data("WallenZD_2022.tse")

 # Assuming tse is a TreeSummarizedExperiment without a rowTree
tse_with_tree <- AddPhyloTree_to_mpa_tse(data.tse = WallenZD_2022.tse, chocophlan_datestamp = "201901", multipleTrees = 1)
# also allowed, because there are 2 trees available with the same timestamp
# tse_with_tree <- AddPhyloTree_to_mpa_tse(WallenZD_2022.tse, chocophlan_datestamp = "201901", multipleTrees = 2)

# this is because we know that metaphlan 3.0 was used for these data:
# https://doi.org/10.1038/s41467-022-34667-x

rowTree(WallenZD_2022.tse)
rowTree(tse_with_tree)
```

The tree version gets saved in the `metadata` slot:

```{r}
tse_with_tree@metadata$PhyloTree
```

if a custom tree is provided, the function won't try to guess, and will store
the tree's variable name

```{r}
tree.list <- load_mpa_tree("201901", multipleTrees = 1)

myCustomTree <- tree.list$Phylo

tse_with_custom_tree <- AddPhyloTree_to_mpa_tse(WallenZD_2022.tse, mpa.tre = myCustomTree, multipleTrees = 1)

tse_with_custom_tree@metadata$PhyloTree
```

## Post-2021 MetaPhlAn data

```{r}
data("JoS_2022.tse")

# add the phylogenetic Tree exploiting the colData information in 
# parkinsonsMetagenomicData
chocophlan_datestamp <- sapply(strsplit(unique(JoS_2022.tse@colData$db_version), "_"), function(x) x[length(x)])

## add it to the TSE by subsetting it and ensuring UNCLASSIFIED remains
JoS_2022.tse <- AddPhyloTree_to_mpa_tse(
  data.tse = JoS_2022.tse,
  chocophlan_datestamp = chocophlan_datestamp, 
  multipleTrees = 1)

```

# Replace standard MetaPhlAn taxonomy with GTDB alternative

works from `MetaPhlAn 4` because it assumes SGBs are the tested units

```{r}
JoS_metaphlan_GTDB.tse <- tse_replace_mpa_with_GTDB_taxonomy(
  data.tse = JoS_2022.tse, 
  chocophlan_datestamp = chocophlan_datestamp)
```

Taxonomy before replacement:

```{r taxonomy_before_replacement}
rowData(JoS_2022.tse)
```

Taxonomy after replacement:

```{r taxonomy_after_replacement}
rowData(JoS_metaphlan_GTDB.tse)
```

# Session info

```{r session_info}
sessionInfo()
```

