---
title: "Searching Phylogenetically Closest Taxa with ChocoPhlAnTaxonomies"
author:
  - name: "Giacomo Antonello"
    affiliation: "CUNY SPH, ISPH"
    email: "giacomo.antonello@sph.cuny.edu"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('ChocoPhlAnTaxonomies')`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Searching Phylogenetically Closest Taxa with ChocoPhlAnTaxonomies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  message   = FALSE,
  warning   = FALSE
)
```

# Introduction

A common task in microbiome research is to identify taxa that are
phylogenetically close to a target species or SGB (Species-level Genome Bin).
This is useful, for example, when a taxon of interest is absent from a dataset
and you want to find the nearest detectable relatives, or when you want to
explore the ecological or functional neighborhood of a focal taxon in the
phylogenetic tree.

`r BiocStyle::Biocpkg("ChocoPhlAnTaxonomies")` provides two complementary
functions for this purpose:

- `search_nearest_taxa_phylo()` — computes phylogenetic distances from a focal
  SGB to all other tips in a MetaPhlAn/ChocoPhlAn tree, returning the closest 
  SGBs phylogenetically first
- `enrich_search_nearest_taxa_phylo()` — takes the distance vector from the
  above and joins it with a full taxonomy table (mpa - Standard MetaPhlAn or 
  GTDB - Genome Taxonomy DataBase), optionally flagging which SGBs are present 
  in a user-provided `r BiocStyle::Biocpkg("TreeSummarizedExperiment")` object.

This vignette walks through both functions using the bundled `JoS_2022` dataset.

# Setup

```{r load-packages}
library(ChocoPhlAnTaxonomies)
library(TreeSummarizedExperiment)
```

## Load example data

We use the `JoS_2022.tse` dataset bundled with the package, which contains
metagenomic profiles from the Jordan et al. 2022 cohort. We attach the
ChocoPhlAn `202403` phylogenetic tree to it using `AddPhyloTree_to_mpa_tse()`.

```{r load-data}
data(JoS_2022.tse)
JoS_2022.tse <- AddPhyloTree_to_mpa_tse(JoS_2022.tse, chocophlan_datestamp = "202403")
JoS_2022.tse
```

Our focal SGB for the examples below is `t__SGB12137`.

```{r define-target}
SGB_target <- "t__SGB12137"
```

# Computing Phylogenetic Distances: `search_nearest_taxa_phylo()`

## Using the tree embedded in the TSE

When you have a `TreeSummarizedExperiment` with an attached phylogenetic tree,
pass `rowTree()` directly. The distances returned are relative to the full set
of tips in that tree — in this case the SGBs detected in the `JoS_2022`
dataset. Because the tree is a subset of the full ChocoPhlAn tree, all SGBs in
the result will, by definition, be present in the dataset.

```{r dist-rowtree}
dist_rowtree.vec <- search_nearest_taxa_phylo(SGB_target, rowTree(JoS_2022.tse))
head(dist_rowtree.vec)
length(dist_rowtree.vec)
```

The result is a named numeric vector sorted by increasing phylogenetic distance.
The self-distance (always 0) is removed, so the vector has one less element
than the number of tips.

## Using the full ChocoPhlAn reference tree

Passing the full reference tree gives distances to all SGBs in a ChocoPhlAn 
phylogenetic tree, most of which will not be detected in any single dataset.
This is the more informative search when you want to explore the full
phylogenetic neighbourhood of a taxon.

```{r load-full-tree}
phyloTree.list <- load_mpa_tree("202403")
```

```{r dist-full-tree}
dist_fulltree.vec <- search_nearest_taxa_phylo(SGB_target, phyloTree.list$Phylo)
head(dist_fulltree.vec)
length(dist_fulltree.vec)
```

## The `t__` prefix is handled automatically

You can pass the SGB name with or without the `t__` prefix — the function adds
it if missing:

```{r no-prefix}
dist_noprefix.vec <- search_nearest_taxa_phylo("SGB12137", phyloTree.list$Phylo)
identical(dist_fulltree.vec, dist_noprefix.vec)
```

# Enriching Results with Taxonomy: `enrich_search_nearest_taxa_phylo()`

`enrich_search_nearest_taxa_phylo()` joins the distance vector with a full
taxonomy table and returns a `data.frame` ready for inspection or export. Two
taxonomy systems are supported: `"mpa"` (MetaPhlAn-style, includes eukaryotes)
and `"GTDB"` (prokaryote-focused, GTDB-Tk style).

## Without a TSE — taxonomy and distances only

### mpa taxonomy

```{r enrich-mpa-notse}
result_mpa <- enrich_search_nearest_taxa_phylo(
  dist.vec              = dist_fulltree.vec,
  taxonomy              = "mpa",
  chocophlan_datestamp  = "202403"
)
head(result_mpa)
```

### GTDB taxonomy

```{r enrich-gtdb-notse}
result_gtdb <- enrich_search_nearest_taxa_phylo(
  dist.vec              = dist_fulltree.vec,
  taxonomy              = "GTDB",
  chocophlan_datestamp  = "202403"
)
head(result_gtdb)
```

Note that GTDB does not contain eukaryote taxonomy. If your tree includes
eukaryotic SGBs, a warning will be raised listing those that could not be
matched — this is expected behaviour, not an error. Use `taxonomy = "mpa"` if
eukaryotes are relevant to your analysis.

## With a TSE — flagging which SGBs are present in your data

Passing a `TreeSummarizedExperiment` object via the `tse` argument adds a
logical column indicating whether each SGB in the result is detected in your
dataset. This makes it straightforward to distinguish between "phylogenetically
close and present" vs "phylogenetically close but absent" taxa.

### Using the full reference tree + TSE

Here, most SGBs will be absent from the dataset (the full tree has many more
SGBs than any single study).

```{r enrich-gtdb-tse-full}
result_gtdb_tse <- enrich_search_nearest_taxa_phylo(
  dist.vec             = dist_fulltree.vec,
  taxonomy             = "GTDB",
  chocophlan_datestamp = "202403",
  tse                  = JoS_2022.tse
)
head(result_gtdb_tse)
```

The last column is named after the `tse` object you passed (here
`present_in_JoS_2022.tse`) and contains `TRUE` if the SGB is in
`rowData(tse)$SGB`, `FALSE` otherwise.

```{r presence-summary}
table(result_gtdb_tse[[ncol(result_gtdb_tse)]])
```

### Using the TSE's own tree + TSE

When the tree comes from `rowTree(tse)`, all SGBs in the result are present in
the dataset by construction:

```{r enrich-gtdb-tse-rowtree}
result_gtdb_tse_rowtree <- enrich_search_nearest_taxa_phylo(
  dist.vec             = dist_rowtree.vec,
  taxonomy             = "GTDB",
  chocophlan_datestamp = "202403",
  tse                  = JoS_2022.tse
)
head(result_gtdb_tse_rowtree)

# All TRUE, as expected
table(result_gtdb_tse_rowtree[[ncol(result_gtdb_tse_rowtree)]])
```

# Practical Tips

**Choosing between the TSE tree and the full reference tree.** Use `rowTree(tse)`
when you want to rank the taxa already present in your dataset by proximity to
the target. Use the full reference tree when you want to find the closest known
relatives regardless of detectability in your study.

**Choosing between mpa and GTDB taxonomy.** mpa taxonomy covers both
prokaryotes and eukaryotes in the ChocoPhlAn database and uses the `k__`/`s__` 
prefix convention. GTDB covers only prokaryotes but provides a more up-to-date 
phylogenomic classification using the `d__`/`s__` convention. If your target SGB 
is a bacterium or archaeon and you want the most current taxonomy, use `"GTDB"`. 
If eukaryotes are in your tree or dataset, use `"mpa"`.

**Missing SGBs in the taxonomy.** A warning listing unmatched SGBs is raised
when some tree tips have no entry in the selected taxonomy file. These SGBs are
silently dropped from the output `data.frame`. If you need to retain them, see 
how many were dropped by reading the warning, and inspect 
`setdiff(names(dist.vec), result$SGB)` after the call.

# Session Info {.unnumbered}

```{r session-info}
sessionInfo()
```
